---
title: "giottoMulti: multi-sample container"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{giottoMulti: multi-sample container}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

## What it is

`giottoMulti` is a container for **multiple `giotto` objects analyzed in a shared expression space**. Each child keeps its own spatial information (different physical / embedding spaces); shared analyses (joint normalization, dim reduction, NN graphs, clustering, integration) live on the parent.

Two design properties worth knowing up front:

- **Federated, not monolithic.** Each child can carry its own on-disk source via `@source`; the multi adds one parent-level source for cross-sample joint artifacts. Samples can be archived, versioned, or handed off independently — the whole multi doesn't have to live in one container.
- **Per-sample analysis surface.** Every child is a full `giotto` object usable standalone. Per-sample QC, normalization, clustering, etc. pre-integration uses the same API as single-sample analysis. The multi adds the joint layer; it doesn't replace the per-sample API.

This is distinct from `joinGiottoObjects()`, which merges multiple giottos into one — destructive, loses per-sample addressability.

## When to use it

Use `giottoMulti` when you have several spatial datasets (e.g. Visium slides, MERFISH samples) that share a feature panel and you want to:

* run joint expression-space analysis across them
* address each sample independently for per-sample operations (transforms, plotting, QC)
* iterate on subsets / filters without re-running integration each time
* keep the option of merging into a single `giotto` later via `joinGiottoObjects()`

## Construction

```{r}
library(GiottoClass)

# Two minimal samples, same feature panel
.mk <- function(ncell, nfeat) {
    m <- matrix(0, nrow = nfeat, ncol = ncell)
    rownames(m) <- paste0("f", seq_len(nfeat))
    colnames(m) <- paste0("c", seq_len(ncell))
    createGiottoObject(expression = m, verbose = FALSE)
}
g1 <- .mk(5, 4)
g2 <- .mk(3, 4)

mg <- createGiottoMulti(list(visium = g1, viz = g2))
mg
```

The `show` output surfaces:

* the children and their cell / feature counts
* the current view: how many cells / features are visible vs total
* which joint shared slots are populated

## Global vs local IDs

Children's cell IDs are *local* (`c1`, `c2`, …). The multi namespaces them as **`sample::id`** globals via `@id_map`:

```{r}
spatIDs(mg)
spatIDs(mg, local = TRUE)
spatIDs(mg, object = "visium")
```

Feature IDs are passthrough by default — the feature panel is shared vocabulary across children.

### Sample identity as a column

When pulling values from the multi via `spatValues()`, sample identity is
surfaced as a `list_ID` column alongside `cell_ID` — handy for downstream
grouping / faceting without re-parsing the namespaced ID:

```{r}
# Pull a feature across all samples
sv <- spatValues(mg, feats = "f1")
sv

# list_ID lets you group / split / color by sample
table(sv$list_ID)
```

This column is added only when `gobject` is a `giottoMulti`; on a plain
`giotto`, `spatValues()` returns just `cell_ID + feats`. The same applies
to multi-feat fetches and metadata-column fetches: any `spatValues()` call
on a gmulti carries `list_ID`.

## Joint expression: assembly then override

`getExpression(mg)` has two modes depending on whether `mg@expression` has been written to:

- **Empty (default after construction)** → assembles a baseline view live from the children: each child's matching `exprObj`, columns renamed to `sample::id`, features intersected, then `cbind`. Useful for QC and inspection before any batch correction has been run.
- **Populated** → reads directly from `mg@expression`. The slot is authoritative; the assembly path is no longer used.

```{r}
# Empty: assembled from children
e <- getExpression(mg)
dim(e[])
colnames(e[])
```

The same destructive-at-first-write lifecycle applies to `@cell_metadata` and `@feat_metadata`: assembled live while empty, materialized on the first `setCellMetadata` / `setFeatureMetadata` call, after which child edits no longer propagate to the multi's view.

After running integration (Harmony, scVI, Symphony, etc.) on the assembled matrix, write the corrected joint matrix back via `setExpression(mg, joint_matrix)` — that's the transition into "populated" mode.

## Subset

`subset()` narrows the joint analysis view of the multi — eagerly: `@id_map` and every *already-populated* joint slot are trimmed in place. Empty joint slots stay empty (subset is a free op on them; assembly later intersects with the narrower `@id_map` automatically). Children (the spatial axis) are not touched.

```{r}
mg_qc <- subset(mg, cells = c("visium::c1", "visium::c2", "viz::c1"))
mg_qc

# Reads reflect the narrowed view
ncol(getExpression(mg_qc)[])

# Children are untouched — their full spatial state is still there
length(spatIDs(mg_qc@objects$visium))
```

To widen back to the full view, use the pre-subset object — `subset()` returns a new `giottoMulti`, so the original `mg` is unchanged:

```{r}
# original is untouched (R copy-on-modify)
spatIDs(mg)
```

If you need a child's spatial state narrowed to a current view, do that on the child explicitly via `@id_map` lookups.

## Per-child operations

Spatial-domain accessors take an `object =` argument to scope per child. Children are returned as-is — `subset()` on the multi narrows the joint analysis view, not children's spatial state.

```{r}
# Spatial locations for one specific child
getSpatialLocations(mg, object = "visium")

# All children — returns a named list
sl_all <- getSpatialLocations(mg)
names(sl_all)
```

Per-child setters require an explicit `object =` target:

```{r, eval=FALSE}
setSpatialLocations(mg, x = my_locs, object = "visium")
setPolygonInfo(mg, x = my_polys, object = "viz")
```

## Slicing a smaller multi

To carve out a giottoMulti with a subset of the children, slice with `[`:

```{r}
mg_one <- mg["visium"]
names(mg_one)
spatIDs(mg_one)
```

There is no stateful "active" pointer to flip — every per-child operation takes an explicit `object =`, so scope is always at the call site. For one-off per-child queries, prefer `spatIDs(mg, object = "visium")` over `spatIDs(mg["visium"])`.

> **Side note: single-object wrap.** `as(g, "giottoMulti")` will coerce a single `giotto` into a one-child `giottoMulti`. This is mostly here so downstream code that branches on `giotto` vs `giottoMulti` can operate uniformly when handed a single sample. For interactive use on one sample, just work with the `giotto` directly.

## Realigning after manual edits

If you edit a child directly (e.g. by replacing it with `mg[["a"]] <- new_g` or by mutating a slot outside the approved op surface), the cached `@id_map` may be out of sync. `initialize(mg)` re-detects via the cached length signature and rebuilds when needed — cheap on every call, no-op when nothing changed.

For an exotic same-length-different-content child edit, just reconstruct: `createGiottoMulti(mg@objects, source = mg@source)` starts fresh, no cached signature.

## Filtering and reframing without mutation

To narrow which cells / features a downstream analysis sees, or to apply
coordinate-frame transforms across children, use `giottoView` and
`giottoSpace` recipes. Both compose naturally with `giottoMulti`:

* A view's `selectSamples()` step picks which children participate.
* A space binds transforms to named samples for cross-sample layouts
  (e.g. positioning visium_a and visium_b side-by-side).
* `materialize(mg, view, space)` runs the recipe once across the multi
  and reuses the survivor-cache across downstream getter calls.

See `vignette("view_and_space", package = "GiottoClass")` for the full
walkthrough.

## Analysis and plotting

The joint analysis layer (normalize / runPCA / runUMAP / clustering /
integration) lives in the `Giotto` analysis package, not here. Those
generics operate on the assembled expression matrix and write the
corrected / reduced result back to the multi's joint slots
(`@expression`, `@dimension_reduction`), at which point downstream reads
flip from "assembled live" to "joint slot is authoritative" (see "Joint
expression" above).

Plotting lives in `GiottoVisuals` and falls into two patterns when handed
a `giottoMulti`:

* **Spatial plots** (`spatPlot2D`, `spatInSituPlotPoints`, ...) dispatch
  per-child via cowplot panels. Pass `space = ":all:"` (the default) for
  one panel per sample, or `space = c("visium_a", ...)` for a subset.
  Cross-sample defined-space panels (multiple children in one panel under
  a shared coord frame) require a `giottoSpace`-based flatten layer that
  is still under development.
* **Non-spatial plots** (`dotPlot`, `violinPlot`, `dimPlot2D`, `plotUMAP`,
  heatmap family) pull data via gmulti-aware getters (`spatValues`,
  `getCellMetadata`, `getDimReduction`) and render one unified plot
  across all samples. Use `cell_color = "list_ID"` to color points by
  sample of origin.

## `giottoMulti` or `joinGiottoObjects()`?

Both put several samples into one workflow, but they differ in what they keep.

| | `giottoMulti` | `joinGiottoObjects()` |
|---|---|---|
| Per-sample addressability | ✅ `samples =` / `"sample::name"` | ❌ merged into one ID space |
| Per-sample spatial state | ✅ each child keeps its own frame | merged with coordinate offsets |
| Federated on-disk per sample | ✅ via `@source` | ❌ merged |
| Per-sample standalone analysis | ✅ each child is a full `giotto` | ❌ |
| Subset semantics | narrows the joint axis; original untouched (value semantics) | destructive |
| Joint analysis layer | ✅ joint slots | inherent — there is only one object |

They are siblings, not nested: one preserves per-sample spaces, the other merges them.
Pick `giottoMulti` when per-sample identity stays load-bearing through the analysis —
QC, normalization, integration, faceted plots, cohort comparisons. Pick
`joinGiottoObjects()` when the samples genuinely become one specimen and you want a
single coordinate space. Convert between them when needed.
